---
title: "BU Physical Therapy Data Analysis"
output:
  pdf_document: default
  html_document: default
date: "March 5, 2018"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyr)
library(plyr)
library(dplyr)
library(Rmisc)
library(knitr)
library(reshape2)
library(gridExtra)
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
## read the data
old <- read.csv("OLD LEFS Data for 2018 ROMS Conf DEIDENTIFIED.csv", header = TRUE, sep = ",")
new <- read.csv("ROMS Data 2018 Conf DEIDENTIFIED.csv", header = TRUE, sep = ",")
```

## Introduction and Summary of Preliminary Analysis

Our clients, James Camarinos & Lee Marinko, have provded patient level outcome data from the BU Physical Therapy and Athletic Training Center. Outcomes are measured in several ways: the difference in the intake versus discharge pain score, the difference in the intake versus discharge level of disability measurement (which is based on a survey depending on the patients' injury type), and the length of stay. The data can be separated into 2 cohorts (one from 2014-2015 and one from 2016-2017). The distinguishing feature between the cohorts is that patients in the more recent cohort (cohort 2) were measured more frequently than patients in the first cohort on pain and disability (at each visit versus sporadically). The first question we address is whether patient outcomes are different across cohorts. This investigation seeks to guage whether the increased level of data collection for the second cohort is associated with different outcomes. Second, using the data from cohort 2, we investigate whether outcomes differ depending on certain factors including gender, age, and body region of injury. 


```{r, echo=FALSE, warning=FALSE, message=FALSE}
## Read data for Q2

#unique(new$Survey)
cohort1 <- old[,c(3,4,6,7,10,11)]
cohort1$Outcome.Diff <- cohort1$LEFS..discharge.-cohort1$LEFS..initial.
cohort1$Pain.diff <- cohort1$Pain.Discharge..worst.-cohort1$Pain.Inital..worst.
cohort1$cohort <- 1
names(cohort1) <- c("LEFS.initial","LEFS.discharge","Pain.initial","Pain.discharge",
                    "Length.of.stay.days", "num.visit", "LEFS.diff","Pain.diff","cohort")
cohort2 <- subset(new,Survey == "LEFS")[,c(14,15,16,17,19,21,20,22)]
cohort2$Pain.diff <- -cohort2$Pain.diff
cohort2$Outcome.Diff <- -cohort2$Outcome.Diff
cohort2$cohort <- 2
names(cohort2) <- c("Pain.initial","LEFS.initial","Pain.discharge","LEFS.discharge",
                    "LEFS.diff", "num.visit", "Pain.diff","Length.of.stay.days","cohort")
data <- na.omit(rbind(cohort1,cohort2))
data <- subset(data, Length.of.stay.days > 0 & num.visit < 100)
```

#### Question 1 - Are Patient Outcomes Different Across Cohorts?

First we visualize the distributions of each raw response variable for cohort 1 and cohort 2 using histograms and boxplots. Note that the cohort 1 sample consists of 112 patients and the cohort 2 sample consists of 671 patients. Since the only survey represented in cohort 1 is the LEFS, we have restricted cohort 2 to LEFS patients. 

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
Length1 <- data.frame(subset(data,cohort == 1)$Length.of.stay.days)
Length1$cohort <- 1
names(Length1) <- c("Length.of.stay.days","Cohort")
Length2 <- data.frame(subset(data,cohort == 2)$Length.of.stay.days)
Length2$cohort <- 2
names(Length2) <- c("Length.of.stay.days","Cohort")
Length <- rbind(Length1, Length2)

visit1 <- data.frame(subset(data,cohort == 1)$num.visit)
visit1$cohort <- 1
names(visit1) <- c("num.visit","Cohort")
visit2 <- data.frame(subset(data,cohort == 2)$num.visit)
visit2$cohort <- 2
names(visit2) <- c("num.visit","Cohort")
visit <- rbind(visit1, visit2)

LEFS.diff1 <- data.frame(subset(data,cohort == 1)$LEFS.diff)
LEFS.diff1$cohort <- 1
names(LEFS.diff1) <- c("LEFS.diff","Cohort")
LEFS.diff2 <- data.frame(subset(data,cohort == 2)$LEFS.diff)
LEFS.diff2$cohort <- 2
names(LEFS.diff2) <- c("LEFS.diff","Cohort")
LEFS.diff <- rbind(LEFS.diff1, LEFS.diff2)

Pain.diff1 <- data.frame(subset(data,cohort == 1)$Pain.diff)
Pain.diff1$cohort <- 1
names(Pain.diff1) <- c("Pain.diff","Cohort")
Pain.diff2 <- data.frame(subset(data,cohort == 2)$Pain.diff)
Pain.diff2$cohort <- 2
names(Pain.diff2) <- c("Pain.diff","Cohort")
Pain.diff <- rbind(Pain.diff1, Pain.diff2)

p1 <- ggplot(Length, aes(x=Length.of.stay.days, fill= factor(Cohort))) + 
  geom_histogram(aes(y=..density..)) + 
  facet_grid(Cohort ~ .)+
  theme(panel.background = element_blank(),
        axis.line = element_line(colour = "black"),legend.position="none") +
  guides(fill=FALSE)
p2 <- ggplot(visit, aes(x=num.visit, fill= factor(Cohort))) + 
  geom_histogram(aes(y=..density..)) +  
  facet_grid(Cohort ~ .)+
  theme(panel.background = element_blank(),
        axis.line = element_line(colour = "black"),legend.position="none") +
  guides(fill=FALSE)
p3 <- ggplot(LEFS.diff, aes(x=LEFS.diff, fill= factor(Cohort))) + 
  geom_histogram(aes(y=..density..)) + 
  facet_grid(Cohort ~ .)+
  theme(panel.background = element_blank(),
        axis.line = element_line(colour = "black"),legend.position="none") +
  guides(fill=FALSE)
p4 <- ggplot(Pain.diff, aes(x=Pain.diff, fill= factor(Cohort))) +
  geom_histogram(aes(y=..density..)) + 
  facet_grid(Cohort ~ .)+
  theme(panel.background = element_blank(),
        axis.line = element_line(colour = "black"),legend.position="none") +
  guides(fill=FALSE)             
             
gridExtra::grid.arrange(p1,p2,p3,p4,ncol=2)
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
rawstat <- data.frame(matrix(ncol = 3,nrow = 4))
names(rawstat) <- c("mean","cohort1","cohort2")
rawstat$mean <- c("LEFS.diff","Pain.diff","Length.of.stay.days","num.visit")
rawstat$cohort1 <- c(mean(subset(data,cohort == 1)$LEFS.diff), 
                     mean(subset(data,cohort == 1)$Pain.diff),
                     mean(subset(data,cohort == 1)$Length.of.stay.days),
                     mean(subset(data,cohort == 1)$num.visit))
rawstat$cohort2 <- c(mean(subset(data,cohort == 2)$LEFS.diff), 
                     mean(subset(data,cohort == 2)$Pain.diff),
                     mean(subset(data,cohort == 2)$Length.of.stay.days),
                     mean(subset(data,cohort == 2)$num.visit))
rawstat = rawstat %>% gather(cohort, ave, cohort1:cohort2)
#rawstat <- rawstat %>% gather(`cohort1`, `cohort2`, key = "cohort", value = "mean")
names(rawstat) <- c("item","cohort","mean")
rawstat$item <- factor(rawstat$item, levels = c("LEFS.diff","Pain.diff","Length.of.stay.days","num.visit"))
```

```{r echo=FALSE, fig.width=12, fig.height=5}
## box plot to show the outliers
gridExtra::grid.arrange(
ggplot(data, aes(x=factor(cohort), y=LEFS.diff, 
                 group = factor(cohort), color = factor(cohort))) + 
  geom_boxplot() + xlab("cohort") +
  theme(panel.background = element_blank(),
        axis.line = element_line(colour = "black"),legend.position="none") +
  guides(fill=FALSE),
ggplot(data, aes(x=factor(cohort), y=Pain.diff, 
                 group = factor(cohort), color = factor(cohort))) + 
  geom_boxplot() + xlab("cohort") +
  theme(panel.background = element_blank(), 
        axis.line = element_line(colour = "black"),legend.position="none") +
  guides(fill=FALSE),
ggplot(data, aes(x=factor(cohort), y=Length.of.stay.days, 
                 group = factor(cohort), color = factor(cohort))) + 
  geom_boxplot() + xlab("cohort") +
  theme(panel.background = element_blank(),
        axis.line = element_line(colour = "black"),legend.position="none") +
  guides(fill=FALSE),
ggplot(data, aes(x=factor(cohort), y=num.visit, 
                 group = factor(cohort), color = factor(cohort))) + 
  geom_boxplot() + xlab("cohort") +
  theme(panel.background = element_blank(),
        axis.line = element_line(colour = "black"),legend.position="none") +
  guides(fill=FALSE),
ncol=4)
```

Below, we visualize the mean and standard error for each of the raw responses. We can see that the average decrease in pain is lower for cohort 2 as compared with cohort 1. This observation is validated in the t-tests below. 


```{r echo=FALSE, fig.width=12, fig.height=5}
## errorbar
LEFS.diff <- summarySE(data, measurevar="LEFS.diff", groupvars="cohort")
Pain.diff <- summarySE(data, measurevar="Pain.diff", groupvars="cohort")
Length.of.stay.days <- summarySE(data, measurevar="Length.of.stay.days", 
                                 groupvars="cohort")
num.visit <- summarySE(data, measurevar="num.visit", 
                                 groupvars="cohort")
gridExtra::grid.arrange(
ggplot(LEFS.diff, aes(x=factor(cohort), y=LEFS.diff,color = factor(cohort)))+
  geom_errorbar(aes(ymin=LEFS.diff-1.96*se, ymax=LEFS.diff+1.96*se), width=.1,size = 1)+
  geom_point(size = 2)+
  theme(legend.position="none")+xlab("cohort") +
  theme(panel.background = element_blank(),
        axis.line = element_line(colour = "black")) +
  guides(fill=FALSE)+
  ggtitle("Error Bars (1.96*SE)") ,
ggplot(Pain.diff, aes(x=factor(cohort), y=Pain.diff,color = factor(cohort)))+
  geom_errorbar(aes(ymin=Pain.diff-1.96*se, ymax=Pain.diff+1.96*se), width=.1,size = 1)+
  geom_point(size = 2)+
  theme(legend.position="none")+xlab("cohort") +
  theme(panel.background = element_blank(),
        axis.line = element_line(colour = "black")) +
  guides(fill=FALSE),
ggplot(Length.of.stay.days, aes(x=factor(cohort), y=Length.of.stay.days,
                                color = factor(cohort)))+
  geom_errorbar(aes(ymin=Length.of.stay.days-1.96*se, 
                    ymax=Length.of.stay.days+1.96*se), width=.1,size = 1)+
  geom_point(size = 2)+
  theme(legend.position="none")+xlab("cohort") +
  theme(panel.background = element_blank(),
        axis.line = element_line(colour = "black")) +
  guides(fill=FALSE),
ggplot(num.visit, aes(x=factor(cohort), y=num.visit,
                                color = factor(cohort)))+
  geom_errorbar(aes(ymin=num.visit-1.96*se, 
                    ymax=num.visit+1.96*se), width=.1,size = 1)+
  geom_point(size = 2)+
  theme(legend.position="none")+xlab("cohort") +
  theme(panel.background = element_blank(),
        axis.line = element_line(colour = "black")) +
  guides(fill=FALSE),
ncol=4)
## bar plot
#ggplot(rawstat, mapping = aes(x=item, y = mean)) + 
#  geom_bar(position = "dodge", stat="identity",aes(fill = cohort), width = 0.5)+
#  theme(panel.background = element_blank(),
#        axis.line = element_line(colour = "black"))
```

Next, we utilize the same set of visualizations for the hake score with respect to pain and LEFS. Note the highest possible score for the LEFS is 80. For pain and LEFS, we calculate the hake score as:  

$\text{hake pain}=(\text{discharge pain}-\text{intake pain})/(\text{intake pain})$  
$\text{hake LEFS}=(\text{discharge LEFS}-\text{intake LEFS})/(80-\text{intake LEFS})$


```{r, echo=FALSE, warning=FALSE, message=FALSE}
## remove patients who had perfect pain or LEFS scores at intake
data = data[-which(data$Pain.initial == 0 | data$LEFS.initial == 80),]


## Calculate Hake score
data$LEFS.hake <- (data$LEFS.discharge - data$LEFS.initial) / (80.1 - data$LEFS.initial)
#We use 80.1 instead of 80 for the maximum score because the denominator cannot be 0
data$Pain.hake <- (data$Pain.discharge - data$Pain.initial) / (data$Pain.initial)

data = data[-which(data$LEFS.hake < -10),] #remove outliers
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
hakestat <- data.frame(matrix(ncol = 3,nrow = 2))
names(hakestat) <- c("mean","cohort1","cohort2")
hakestat$mean <- c("LEFS.diff.hake","Pain.diff.hake")



hakestat$cohort1 <- c(mean(subset(data,cohort == 1)$LEFS.hake), 
                      mean(subset(data,cohort == 1)$Pain.hake))
hakestat$cohort2 <- c(mean(subset(data,cohort == 2)$LEFS.hake), 
                      mean(subset(data,cohort == 2)$Pain.hake))

hakestat = hakestat %>% gather(cohort, ave, cohort1:cohort2)

#hakestat <- hakestat %>% 
#  gather(`cohort1`, `cohort2`, key = "cohort", value = "mean")
names(hakestat) <- c("item","cohort","mean")
hakestat$item <- factor(hakestat$item, levels = c("LEFS.diff.hake","Pain.diff.hake"))
```

```{r echo=FALSE, fig.width=12, fig.height=5}
## box plot to show the outliers
gridExtra::grid.arrange(
ggplot(data, aes(x=factor(cohort), y=LEFS.hake, 
                 group = factor(cohort), color = factor(cohort))) + 
  geom_boxplot() + xlab("cohort") +
  theme(panel.background = element_blank(),
        axis.line = element_line(colour = "black"),legend.position="none") +
  guides(fill=FALSE),
ggplot(data, aes(x=factor(cohort), y=Pain.hake, 
                 group = factor(cohort), color = factor(cohort))) + 
  geom_boxplot() + xlab("cohort") +
  theme(panel.background = element_blank(),
        axis.line = element_line(colour = "black"),legend.position="none") +
  guides(fill=FALSE),
ncol=2)
## errorbar
LEFS.hake <- summarySE(data, measurevar="LEFS.hake", groupvars="cohort")
Pain.hake <- summarySE(data, measurevar="Pain.hake", groupvars="cohort")
gridExtra::grid.arrange(
ggplot(LEFS.hake, aes(x=factor(cohort), y=LEFS.hake,color = factor(cohort)))+
  geom_errorbar(aes(ymin=LEFS.hake-1.96*se, ymax=LEFS.hake+1.96*se), width=.1,size=1)+
  geom_point(size=2)+
  theme(legend.position="none")+xlab("cohort") +
  theme(panel.background = element_blank(),axis.line = element_line(colour = "black")) +
  guides(fill=FALSE),
ggplot(Pain.hake, aes(x=factor(cohort), y=Pain.hake,color = factor(cohort)))+
  geom_errorbar(aes(ymin=Pain.hake-1.96*se, ymax=Pain.hake+1.96*se), width=.1,size=1)+
  geom_point(size=2)+
  theme(legend.position="none")+xlab("cohort") +
  theme(panel.background = element_blank(),axis.line = element_line(colour = "black")) +
  guides(fill=FALSE),
ncol=2)
## bar plot
#ggplot(hakestat, mapping = aes(x=item, y = mean)) + 
#  geom_bar(position = "dodge", stat="identity",aes(fill = cohort))+
#  theme(panel.background = element_blank(),
#        axis.line = element_line(colour = "black"))
```


##### Statistical Tests

We carry out t-tests for each response variable to test the null hypothesis that the mean response for cohort 1 is equal to the mean response for cohort 2. The first table shows the results of the t-test for the raw outcome scores whereas the second table shows the results of the t-tests for the normalized hake scores.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# t-test
t1 = t.test(Length.of.stay.days ~ cohort, data = data, paired = FALSE, conf.level = 0.95)
t1_ci = paste("(",round(t1$conf.int[1],2),",",round(t1$conf.int[2],2),")", sep="")
t2 = t.test(num.visit ~ cohort, data = data, paired = FALSE, conf.level = 0.95)
t2_ci = paste("(",round(t2$conf.int[1],2),",",round(t2$conf.int[2],2),")", sep="")
t3 = t.test(Pain.diff ~ cohort, data = data, paired = FALSE, conf.level = 0.95)
t3_ci = paste("(",round(t3$conf.int[1],2),",",round(t3$conf.int[2],2),")", sep="")
t4 = t.test(LEFS.diff ~ cohort, data = data, paired = FALSE, conf.level = 0.95)
t4_ci = paste("(",round(t4$conf.int[1],2),",",round(t4$conf.int[2],2),")", sep="")

group <- c("Length.of.stay.days", "num.visit", "Pain.diff", "LEFS.diff")
CI <- c(t1_ci, t2_ci, t3_ci, t4_ci)
rejection <- c("No", "Yes", "Yes", "No")
summary_table <- t(rbind(group, CI, rejection))
summary_table <- data.frame(summary_table)
names(summary_table) = c("Construct", "Confidence interval for diff.","Reject H0?")
kable(summary_table)
#kable(rename(summary_table, c("group" = "Construct", "CI" = "Confidence Interval for difference",
#"rejection" = "Reject $H_0$: true difference in means is equal to 0?")))
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
## t-test
t1 = t.test(Pain.hake ~ cohort, data = data, paired = FALSE, conf.level = 0.95)
t1_ci = paste("(",round(t1$conf.int[1],2),",",round(t1$conf.int[2],2),")", sep="")
t2 = t.test(LEFS.hake ~ cohort, data = data, paired = FALSE, conf.level = 0.95)
t2_ci = paste("(",round(t2$conf.int[1],2),",",round(t2$conf.int[2],2),")", sep="")

group <- c("Pain.hake", "LEFS.hake")
CI <- c(t1_ci, t2_ci)
rejection <- c("Yes", "No")
summary_table <- t(rbind(group, CI, rejection))
summary_table <- data.frame(summary_table)
names(summary_table) = c("Response", "Confidence interval for diff.","Reject H0?")
kable(summary_table)

#kable(rename(summary_table, c("group" = "Construct", "CI" = "Confidence Interval for difference", "rejection" = "Reject $H_0$:true #difference in means equals to 0?")))
```


#### Question 2 - Do Patient Outcomes Depend on Gender/ Age/ Body Region?

Next, we investigate whether outcomes depend on gender, age, and body region. Before fitting linear regression models, we visually explore the bivariate relationships between pain difference and the explanatory variables - age, survey type, and gender. We can see that there does not appear to be any discernable difference in this particular response across the range of the explanatory variables. This lack of effect is further validated in the regression models fit below.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
diff_m <- new[,c(2,4,6,7,14,15,19,20,21,22)]
diff_m <- na.omit(diff_m)
diff_m$Pain.diff <- -1* diff_m$Pain.diff
diff_m$Outcome.Diff <- -1* diff_m$Outcome.Diff


colnames(diff_m) <- c("body","gender","age","survey","admit_pain","admit_outcome","outcome","pain","visit","length_stay")

diff_m$body <- as.factor(diff_m$body)
diff_m$gender <- as.factor(diff_m$gender)
diff_m$survey <- as.factor(diff_m$survey)

```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
diff_h_pain <- diff_m
diff_h_pain$HakePain <- diff_h_pain$pain / abs(0-diff_h_pain$admit_pain)

diff_outcome_DASH <- filter(diff_m, survey == "DASH")
diff_outcome_LEFS <- filter(diff_m, survey == "LEFS")
diff_outcome_ODI <- filter(diff_m, survey == "ODI")
diff_outcome_NECK <- filter(diff_m, survey == "NECK")
diff_outcome_KNEE <- filter(diff_m, survey == "KNEE")

diff_h_outcome_DASH <- diff_outcome_DASH
diff_h_outcome_LEFS <- diff_outcome_LEFS
diff_h_outcome_ODI <- diff_outcome_ODI
diff_h_outcome_NECK <- diff_outcome_NECK
diff_h_outcome_KNEE <- diff_outcome_KNEE

diff_h_outcome_DASH$HakeOutcome <- diff_h_outcome_DASH$outcome / abs(0-diff_h_outcome_DASH$admit_outcome)
diff_h_outcome_LEFS$HakeOutcome <- diff_h_outcome_LEFS$outcome / abs(80-diff_h_outcome_LEFS$admit_outcome)
diff_h_outcome_ODI$HakeOutcome <- diff_h_outcome_ODI$outcome / abs(0-diff_h_outcome_ODI$admit_outcome)
diff_h_outcome_NECK$HakeOutcome <- diff_h_outcome_NECK$outcome / abs(0-diff_h_outcome_NECK$admit_outcome)
diff_h_outcome_KNEE$HakeOutcome <- diff_h_outcome_KNEE$outcome / abs(100-diff_h_outcome_KNEE$admit_outcome)

#hake pain
diff_h_pain <- na.omit(diff_h_pain)
diff_h_pain <- diff_h_pain[is.finite(diff_h_pain[,11]),]
#outcome
diff_outcome_DASH <- na.omit(diff_outcome_DASH)
#diff_outcome_DASH <- diff_outcome_DASH[is.finite(diff_outcome_DASH[,11]),]

diff_outcome_LEFS <- na.omit(diff_outcome_LEFS)
#diff_outcome_LEFS <- diff_outcome_LEFS[is.finite(diff_outcome_LEFS[,11]),]

diff_outcome_ODI <- na.omit(diff_outcome_ODI)
#diff_outcome_ODI <- diff_outcome_ODI[is.finite(diff_outcome_ODI[,11]),]

diff_outcome_NECK <- na.omit(diff_outcome_NECK)
#diff_outcome_NECK <- diff_outcome_NECK[is.finite(diff_outcome_NECK[,11]),]

diff_outcome_KNEE <- na.omit(diff_outcome_KNEE)
#diff_outcome_KNEE <- diff_outcome_KNEE[is.finite(diff_outcome_KNEE[,11]),]

#hake outcome
diff_h_outcome_DASH <- na.omit(diff_h_outcome_DASH)
diff_h_outcome_DASH <- diff_h_outcome_DASH[is.finite(diff_h_outcome_DASH[,11]),]

diff_h_outcome_LEFS <- na.omit(diff_h_outcome_LEFS)
diff_h_outcome_LEFS <- diff_h_outcome_LEFS[is.finite(diff_h_outcome_LEFS[,11]),]

diff_h_outcome_ODI <- na.omit(diff_h_outcome_ODI)
diff_h_outcome_ODI <- diff_h_outcome_ODI[is.finite(diff_h_outcome_ODI[,11]),]

diff_h_outcome_NECK <- na.omit(diff_h_outcome_NECK)
diff_h_outcome_NECK <- diff_h_outcome_NECK[is.finite(diff_h_outcome_NECK[,11]),]

diff_h_outcome_KNEE <- na.omit(diff_h_outcome_KNEE)
diff_h_outcome_KNEE <- diff_h_outcome_KNEE[is.finite(diff_h_outcome_KNEE[,11]),]
```


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
p1 = ggplot(diff_m, aes(x=age, y=pain)) + geom_point() + ggtitle("Pain vs. Age")
p2 = ggplot(diff_m, aes(x=survey, y=pain)) + geom_point() + geom_jitter() + ggtitle("Pain vs. Survey")
p3 = ggplot(diff_m, aes(x=gender, y=pain)) + geom_point() + geom_jitter() + ggtitle("Pain vs. Gender")
grid.arrange(p1,p2,p3, nrow=1)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
p1 = ggplot(diff_m, aes(x=age, y=length_stay)) + geom_point() + ggtitle("length of stay vs. Age")
p2 = ggplot(diff_m, aes(x=survey, y=length_stay)) + geom_point() + geom_jitter() + ggtitle("length of stay vs. Survey")
p3 = ggplot(diff_m, aes(x=gender, y=length_stay)) + geom_point() + geom_jitter() + ggtitle("length of stay vs. Gender")
grid.arrange(p1,p2,p3, nrow=1)
```

Next, we fit a series of linear regression models to test whether outcomes depend on gender, age, and body region. We split our models into 2 catgories, depending on whether the response variables are survey specific (change in survey outcome score) or not (pain, length of stay, number of visits). The first set of models we present are not survey specific. 


```{r, echo=FALSE, warning=FALSE, message=FALSE}
#non-survey-based models
pain_m <- lm(pain ~ age + survey + gender + body, data = diff_m)
pain_hake <- lm(HakePain ~ age + survey + gender + body, data = diff_h_pain)
visit_m <- lm(visit ~ age + survey + gender + body, data = diff_m)
length_m <- lm(length_stay ~ age + survey + gender + body, data = diff_m)
#outcome models
outcome_DASH <- lm(outcome ~ age + gender + body, data = diff_outcome_DASH)
outcome_LEFS <- lm(outcome ~ age + gender + body, data = diff_outcome_LEFS)
outcome_ODI <- lm(outcome ~ age + gender + body, data = diff_outcome_ODI)
outcome_NECK <- lm(outcome ~ age + gender + body, data = diff_outcome_NECK)
outcome_KNEE <- lm(outcome ~ age + gender + body, data = diff_outcome_KNEE)
#hake outcome models
outcome_hake_DASH <- lm(HakeOutcome ~ age + gender + body, data = diff_h_outcome_DASH)
outcome_hake_LEFS <- lm(HakeOutcome ~ age + gender + body, data = diff_h_outcome_LEFS)
outcome_hake_ODI <- lm(HakeOutcome ~ age + gender + body, data = diff_h_outcome_ODI)
outcome_hake_NECK <- lm(HakeOutcome ~ age + gender + body, data = diff_h_outcome_NECK)
outcome_hake_KNEE <- lm(HakeOutcome ~ age + gender + body, data = diff_h_outcome_KNEE)
```


```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(pander)
#non-survey-based models
pander(pain_m)
pander(pain_hake)
pander(visit_m)
pander(length_m)
```

Next, for the survey specific models, we fit separate models for each of 5 survey types - first for the raw differences and then for the hake score changes. The survey order of these models is: DASH, LEFS, ODI, NECK, KNEE.

#### DASH score as response
```{r, echo=FALSE, warning=FALSE, message=FALSE}
#outcome models
pander(outcome_DASH)
```

#### LEFS score as response
```{r, echo=FALSE, warning=FALSE, message=FALSE}
pander(outcome_LEFS)
```

#### ODI score as response
```{r, echo=FALSE, warning=FALSE, message=FALSE}
pander(outcome_ODI)
```

#### NECK score as response
```{r, echo=FALSE, warning=FALSE, message=FALSE}
pander(outcome_NECK)
```

#### KNEE score as response
```{r, echo=FALSE, warning=FALSE, message=FALSE}
pander(outcome_KNEE)
```

#### DASH Hake score as response
```{r, echo=FALSE, warning=FALSE, message=FALSE}
#outcome hakes models
pander(outcome_hake_DASH)
```



#### LEFS Hake score as response
```{r, echo=FALSE, warning=FALSE, message=FALSE}
pander(outcome_hake_LEFS)
```

#### ODI Hake score as response
```{r, echo=FALSE, warning=FALSE, message=FALSE}
pander(outcome_hake_ODI)
```

#### NECK Hake score as response
```{r, echo=FALSE, warning=FALSE, message=FALSE}
pander(outcome_hake_NECK)
```

#### KNEE Hake score as response
```{r, echo=FALSE, warning=FALSE, message=FALSE}
pander(outcome_hake_KNEE)
```

`

